---
title: "A Mapping of Denver Marijuana Businesses and Arrests"
output:
  html_document:
    df_print: paged
---
Author: David Martinez</br>
Institution: Regis University
CLass: MSDS692 Data Science Practicum #1</br>
Professor: Dr. Michael Busch

### Introduction</br>
On November 7th, 2000, voters in Colorado amended the state Constitution to allow the sale and use of Marijuana upon written consent by medical professional (Amendment 20, Colorado Constitution, 2000). Twelve years later, Colorado approved the sale and use of marijuana recreational use for adults over the age of twenty-one (Amendment 64, COlorado Constitution, 2012). Denver, the state capital and the largest population center in Colorado, has published data records since 2010 for medical marijuana and 2013 for recreational marijuana, including sales, government revenue, licensing information, and crime statistics.</br>

This project will focus on identifying the types and locations of Marijuana businesses as well as the types and locations of arrests made.</br>

The following datasets are used in this project: </br>
https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-marijuana-active-business-licenses
https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-crime-marijuana

Other references:</br>
https://ballotpedia.org/Marijuana_on_the_ballot</br>
https://developers.google.com/maps/documentation/geocoding/usage-and-billing</br>

citations: 
ggmap - D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, 5(1), 144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf</br>

stringr - Hadley Wickham (2017). stringr: Simple, Consistent Wrappers for Common String Operations. R package version 1.2.0.
  https://CRAN.R-project.org/package=stringr</br>

ggplot2 - H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2009.</br>

dplyr - Hadley Wickham, Romain Francois, Lionel Henry and Kirill Müller (2017). dplyr: A Grammar of Data Manipulation. R package
  version 0.7.4. https://CRAN.R-project.org/package=dplyr</br>

RgoogleMaps - Markus Loecher and Karl Ropkins (2015). RgoogleMaps and loa: Unleashing R Graphics Power on Map Tiles. Journal of
  Statistical Software 63(4), 1-18. URL http://www.jstatsoft.org/v63/i04/.</br>

### Retrieve Denver Marijuana Licenses
```{r message=F}
library(stringr)  #string operations
library(ggplot2)  #graphics
library(ggmap)    #used for geocoding 
library(dplyr)    #dataframe manipulation

#retrieve dataset from denvergov.org
denver_mj_licenses <- read.csv("https://www.denvergov.org/media/gis/DataCatalog/marijuana_active_business_licenses/csv/marijuana_active_business_licenses.csv", header=TRUE, stringsAsFactors = FALSE)

#create factor for license type 
denver_mj_licenses$License.Type <- factor(denver_mj_licenses$License.Type)

#combine levels and rename for easier reading
denver_mj_licenses$License.Type <- plyr::revalue(denver_mj_licenses$License.Type, c("Med Marijuana Inf Prod Manuf" = "Infrastructure Production Manufacturing", "Retail Marijuana Inf Prod Mfg" = "Infrastructure Production Manufacturing", "Med Marijuana Test Facility" = "Marijuana Test Facility", "Retail Marijuana Test Facility" = "Marijuana Test Facility", "Med Marijuana Opt Prem Cultiv" = "Marijuana Grow Center", "Retail Marij Opt. Prem. Cultiv" = "Marijuana Grow Center"))

#create factor for zip code
denver_mj_licenses$Facility.Zip.Code <- factor(denver_mj_licenses$Facility.Zip.Code)

#current license status and expiration date don't appear to be very interesting so they can go
denver_mj_licenses <- denver_mj_licenses[-c(5:6)]

#convert street address information to one field for geocoding
num <- paste(word(denver_mj_licenses$Facility.Street.Number))
dir <- paste(word(denver_mj_licenses$Facility.Pre.Direction))
street <- paste(word(denver_mj_licenses$Facility.Street.Name))
type <- paste(word(denver_mj_licenses$Facility.Street.Type))
denver_mj_licenses$ADDRESS <- paste(num, dir, street, type, ", DENVER, CO", sep=" ")

#remove old address information (except zip code)
denver_mj_licenses <- denver_mj_licenses[-c(5:10)]

#geocode for lat/long - Data Science Toolkit (dsk) is used here instead of to retrieve the lat/long 
for(i in 1:nrow(denver_mj_licenses)) {
  result <- geocode(denver_mj_licenses$ADDRESS[i], output="latlona", source="dsk")
  denver_mj_licenses$LONGITUDE[i] <- as.numeric(result[1])
  denver_mj_licenses$LATITUDE[i] <- as.numeric(result[2])
}

#save for posterity
#write.csv(denver_mj_licenses, "mj_licenses_geocoded.csv", row.names=FALSE)

summary(denver_mj_licenses)

#cleanup again - darn OCD
rm(i, num, dir, street, type, result)

#sort by License.Type
denver_mj_licenses <- within(denver_mj_licenses, License.Type <- factor(License.Type, 
                                                                        levels=names(sort(table(License.Type), decreasing=TRUE))))

#generate plot to show graphically the number of license types issued
ggplot(denver_mj_licenses,aes(x=License.Type, fill=License.Type))+
  geom_bar()+
  labs(title="Count of License Types issued", x= "License Type")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#sort by Facility Zip Code
denver_mj_licenses <- within(denver_mj_licenses, Facility.Zip.Code <- factor(Facility.Zip.Code, 
                                                                      levels=names(sort(table(Facility.Zip.Code), decreasing=TRUE))))
#generate plot to show where licenses are being issued
ggplot(denver_mj_licenses,aes(x=Facility.Zip.Code, fill=License.Type))+
  geom_bar()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#create a dataframe of the types of licenses issued to each zip code  
zip_tab <- as.data.frame(table(denver_mj_licenses$Facility.Zip.Code, denver_mj_licenses$License.Type))

#rename for more descriptive tags
zip_tab <- rename(zip_tab, ZipCode = Var1, License.Type = Var2)

#filter out the 0 frequency occurences
zip_tab <- filter(zip_tab, zip_tab$Freq > 0)

#generate a plot to show the types of licenses issued to each zip code
ggplot(zip_tab,aes(x=ZipCode, fill=License.Type))+
  geom_bar()+
  labs(title="Types of licenses issued by zip code", x= "Denver Zip Codes")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
### Generating Maps




### Clean and save marijuana crime dataset
```{r message=F}

#retrieve the DPD marijuana crime file from Denver open data portal
denver_mj_crime <- read.csv("https://www.denvergov.org/media/gis/DataCatalog/crime_marijuana/csv/crime_marijuana.csv", header=TRUE, strip.white = TRUE, stringsAsFactors = FALSE)

#write original file to disk for posterity
write.csv(denver_mj_crime, "crime_marijuana.csv", row.names=FALSE)

#incident ID to character
denver_mj_crime$INCIDENT_ID <- as.character(denver_mj_crime$INCIDENT_ID)

#Only one date field is needed
denver_mj_crime <- data.frame(denver_mj_crime[-c(2:3, 6:7, 10:11)])

#The Report date needs to be in an R compliant format
day <- paste(word(denver_mj_crime$REPORTDATE, 1, sep="-"))
month <- paste(word(denver_mj_crime$REPORTDATE, 2, sep="-"))
month <- plyr::revalue(month, c("JAN"="01", "FEB"="02", "MAR"="03", "APR"="04", "MAY"="05", "JUN"="06", "JUL"="07", "AUG"="08", "SEP"="09", "OCT"="10", "NOV"="11", "DEC"="12"))
year <- as.integer(paste(word(denver_mj_crime$REPORTDATE, -1, sep="-")))
year <- paste(year+2000)

#denver_mj_crime$REPORTDATE <- paste(year, month, sep="-")
denver_mj_crime$REPORTDATE <- paste(year, month, day, sep="-")
denver_mj_crime$REPORTDATE <- as.Date(denver_mj_crime$REPORTDATE)

#append city and state information to address - necessary for geocoding
denver_mj_crime$INCIDENT_ADDRESS <- sapply(denver_mj_crime$INCIDENT_ADDRESS , paste, ", Denver, CO", sep="")

#geo_x, geo_y are not needed for this activity
#denver_mj_crime <- data.frame(denver_mj_crime[-c(4:5)])

#convert fields to factors
denver_mj_crime$DISTRICT_ID <- factor(denver_mj_crime$DISTRICT_ID)
denver_mj_crime$PRECINCT_ID <- factor(denver_mj_crime$PRECINCT_ID)
denver_mj_crime$OFFENSE_CATEGORY_ID <- factor(denver_mj_crime$OFFENSE_CATEGORY_ID)
denver_mj_crime$MJ_RELATION_TYPE <- factor(denver_mj_crime$MJ_RELATION_TYPE)
denver_mj_crime$NEIGHBORHOOD_ID <- factor(denver_mj_crime$NEIGHBORHOOD_ID)

#geocode to add Longitude/Latitude data
for(i in 1:nrow(denver_mj_crime)) {
  result <- geocode(denver_mj_crime$INCIDENT_ADDRESS[i], output="latlona", source="dsk")
  denver_mj_crime$LONGITUDE[i] <- as.numeric(result[1])
  denver_mj_crime$LATITUDE[i] <- as.numeric(result[2])
}

summary(denver_mj_crime)

#remove intermediate variables and capture the clean file for posterity
rm(result, i, day, month, year)

#sort by neighborhood
denver_mj_crime <- within(denver_mj_crime, NEIGHBORHOOD_ID <- factor(NEIGHBORHOOD_ID, 
                                                                      levels=names(sort(table(NEIGHBORHOOD_ID), decreasing=FALSE))))

#generate plot to show neighborhoods and arrests
ggplot(denver_mj_crime,aes(x=NEIGHBORHOOD_ID, fill=OFFENSE_CATEGORY_ID))+
  geom_bar()+
  coord_flip()+
  labs(title="License Types issued by zip code", x= "Denver Neighborhoods")+
  theme_bw()

#sort by crime type
denver_mj_crime <- within(denver_mj_crime, OFFENSE_CATEGORY_ID <- factor(OFFENSE_CATEGORY_ID, 
                                                                      levels=names(sort(table(OFFENSE_CATEGORY_ID), decreasing=FALSE))))

#generate plot to show where licenses are being issued
ggplot(denver_mj_crime,aes(x=OFFENSE_CATEGORY_ID, fill=OFFENSE_CATEGORY_ID))+
  geom_bar()+
  coord_flip()+
  labs(title="Marijuana Related Arrests Made", x= "Arrest Type")+
  theme_bw()
```

This section requires a Google developers API Key: (https://developers.google.com/maps/documentation/geocoding/usage-and-billing). For security purposes this section is masked.
```{r echo=FALSE}
#API Key is required from google for the next section.
apikey = "AIzaSyDrMbhq1f_Be9omWdC4QlPKczFlazedMlo"
```

```{r}
library(RgoogleMaps) #interface to google maps 

#API key
#apikey <- "your api key"

#generate denver map. 
denver_map <- GetMap(center = c(lat = mean(denver_mj_licenses$LATITUDE), lon = mean(denver_mj_licenses$LONGITUDE)), destfile="denver_map.png", size = c(640, 640), zoom = 12, sensor = "false", maptype = "roadmap", format = "png", RETURNIMAGE = TRUE, GRAYSCALE = FALSE, NEWMAP = TRUE, SCALE = 1, API_console_key = apikey, verbose = 0)

p1 <- PlotOnStaticMap(denver_map, lat = denver_mj_licenses$LATITUDE, lon = denver_mj_licenses$LONGITUDE, destfile="denver_map.png", GRAYSCALE = FALSE, add = FALSE, FUN = points, NEWMAP = TRUE, TrueProj = TRUE, axes = TRUE, atX = NULL, atY = NULL, col = 3, pch=9)
legend("topleft", legend = "Location of Marijuana Businesses", col = "green", bg = "white", pch=9)

p2 <- PlotOnStaticMap(denver_map, lat = denver_mj_crime$LATITUDE, lon = denver_mj_crime$LONGITUDE, destfile="denver_map.png", GRAYSCALE = FALSE, add = FALSE, FUN = points, NEWMAP = TRUE, TrueProj = TRUE, axes = TRUE, atX = NULL, atY = NULL, col = 2, pch=12)
legend("topleft", legend = "Location of Marijuana Arrests", col = "red", bg = "white", pch=12)
```



