---
title: "A Mapping of Denver Marijuana Businesses and Arrests"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---
Author: David Martinez</br>

Institution: Regis University</br>

Class: MSDS692 Data Science Practicum #1</br>

Professor: Dr. Michael Busch</br>


### Introduction</br>
On November 7th, 2000, voters in Colorado amended the state Constitution to allow the sale and use of Marijuana upon written consent by medical professional (Amendment 20, Colorado Constitution, 2000). Twelve years later, Colorado approved the sale and use of marijuana recreational use for adults over the age of twenty-one (Amendment 64, COlorado Constitution, 2012). Denver, the state capital and the largest population center in Colorado, has published data records since 2010 for medical marijuana and 2013 for recreational marijuana, including sales, government revenue, licensing information, and crime statistics.</br>

This project will focus on identifying the types and locations of Marijuana businesses as well as the types and locations of arrests made.</br>

The following datasets are used in this project: </br>

https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-marijuana-active-business-licenses</br>

https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-crime-marijuana</br>

Other references:</br>
https://ballotpedia.org/Marijuana_on_the_ballot</br>

https://developers.google.com/maps/documentation/geocoding/usage-and-billing</br>

## citations: 
ggmap - D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, 5(1), 144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf</br>

stringr - Hadley Wickham (2017). stringr: Simple, Consistent Wrappers for Common String Operations. R package version 1.2.0.
  https://CRAN.R-project.org/package=stringr</br>

ggplot2 - H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2009.</br>

dplyr - Hadley Wickham, Romain Francois, Lionel Henry and Kirill Müller (2017). dplyr: A Grammar of Data Manipulation. R package
  version 0.7.4. https://CRAN.R-project.org/package=dplyr</br>

RgoogleMaps - Markus Loecher and Karl Ropkins (2015). RgoogleMaps and loa: Unleashing R Graphics Power on Map Tiles. Journal of
  Statistical Software 63(4), 1-18. URL http://www.jstatsoft.org/v63/i04/.</br>

## Dependencies
```{r message=FALSE}
#dependencies
library(stringr)  #string operations
library(ggplot2)  #graphics
library(ggmap)    #used for geocoding
library(RgoogleMaps)    #used for mapping lat/long to static map
library(dplyr)    #dataframe manipulation
```

## API Key
A google developers API Key is necessary for this project. Said key can be optained from google via the following link: (https://developers.google.com/maps/documentation/geocoding/usage-and-billing).</br>

This operation is masked to protect my key. The next section contains the code (commented out).<br>

```{r}
#API Key is required from google for the next section.
#register_google(key = 'YOUR KEY HERE')
```

### Retrieve Denver Marijuana Licenses
```{r message=F}

#retrieve dataset from local directory (or you can download it from): #https://github.com/davelovesdata/Denver-Marijuana-Locations/blob/master/mj_licenses_geocoded.csv)
denver_mj_licenses <- read.csv("C:/Users/David/Documents/MSDS692/mj_licenses_geocoded.csv", header=TRUE)

#geocode for lat/long via google 
for(i in 1:nrow(denver_mj_licenses)) {
  result <- geocode(as.character(denver_mj_licenses$ADDRESS[i]), output="latlona", source="google")
  denver_mj_licenses$LONGITUDE[i] <- as.numeric(result[1])
  denver_mj_licenses$LATITUDE[i] <- as.numeric(result[2])
}

summary(denver_mj_licenses)

#sort by License.Type
denver_mj_licenses <- within(denver_mj_licenses, License.Type <- factor(License.Type, 
                                                                        levels=names(sort(table(License.Type), decreasing=TRUE))))

#generate plot to show graphically the number of license types issued
ggplot(denver_mj_licenses,aes(x=License.Type, fill=License.Type))+
  geom_bar()+
  labs(title="Count of License Types issued", x= "License Type")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#sort by Facility Zip Code
denver_mj_licenses <- within(denver_mj_licenses, Facility.Zip.Code <- factor(Facility.Zip.Code, 
                                                                      levels=names(sort(table(Facility.Zip.Code), decreasing=TRUE))))
#generate plot to show where licenses are being issued
ggplot(denver_mj_licenses,aes(x=Facility.Zip.Code, fill=License.Type))+
  geom_bar()+
  labs(title="Licenses issued per Zip Code", x="Zip Code")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#create a dataframe of the types of licenses issued to each zip code  
zip_tab <- as.data.frame(table(denver_mj_licenses$Facility.Zip.Code, denver_mj_licenses$License.Type))

#rename for more descriptive tags
zip_tab <- rename(zip_tab, ZipCode = Var1, License.Type = Var2)

#filter out the 0 frequency occurences
zip_tab <- filter(zip_tab, zip_tab$Freq > 0)

#generate a plot to show the types of licenses issued to each zip code
ggplot(zip_tab,aes(x=ZipCode, fill=License.Type))+
  geom_bar()+
  labs(title="Types of licenses issued by zip code", x= "Denver Zip Codes")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### Clean and save marijuana crime dataset
```{r message=F}

#retrieve dataset from local directory (or you can download it from): #https://github.com/davelovesdata/Denver-Marijuana-Locations/blob/master/mj_crime_geocoded.csv)
denver_mj_crime <- read.csv("C:/Users/David/Documents/MSDS692/mj_crime_geocoded.csv", header=TRUE)

#geocode to add Longitude/Latitude data
for(i in 1:nrow(denver_mj_crime)) {
  result <- geocode(as.character(denver_mj_crime$INCIDENT_ADDRESS[i]), output="latlona", source="google")
  denver_mj_crime$LONGITUDE[i] <- as.numeric(result[1])
  denver_mj_crime$LATITUDE[i] <- as.numeric(result[2])
}

summary(denver_mj_crime)

#remove intermediate variables and capture the clean file for posterity
rm(result, i, day, month, year)

#sort by neighborhood
denver_mj_crime <- within(denver_mj_crime, NEIGHBORHOOD_ID <- factor(NEIGHBORHOOD_ID, 
                                                                      levels=names(sort(table(NEIGHBORHOOD_ID), decreasing=FALSE))))

#generate plot to show neighborhoods and arrests
ggplot(denver_mj_crime,aes(x=NEIGHBORHOOD_ID, fill=OFFENSE_CATEGORY_ID))+
  geom_bar()+
  coord_flip()+
  labs(title="License Types issued by zip code", x= "Denver Neighborhoods")+
  theme_bw()

#sort by crime type
denver_mj_crime <- within(denver_mj_crime, OFFENSE_CATEGORY_ID <- factor(OFFENSE_CATEGORY_ID, 
                                                                      levels=names(sort(table(OFFENSE_CATEGORY_ID), decreasing=FALSE))))

#generate plot to show where licenses are being issued
ggplot(denver_mj_crime,aes(x=OFFENSE_CATEGORY_ID, fill=OFFENSE_CATEGORY_ID))+
  geom_bar()+
  coord_flip()+
  labs(title="Marijuana Related Arrests Made", x= "Arrest Type")+
  theme_bw()
```

```{r}
#apikey <- "YOUR KEY HERE"
```

```{r}
library(RgoogleMaps) #interface to google maps 

#generate denver map. 
denver_map <- GetMap(center = c(lat = mean(denver_mj_licenses$LATITUDE), lon = mean(denver_mj_licenses$LONGITUDE)), destfile="denver_map.png", size = c(640, 640), zoom = 12, sensor = "false", maptype = "roadmap", format = "png", RETURNIMAGE = TRUE, GRAYSCALE = FALSE, NEWMAP = TRUE, SCALE = 1, API_console_key = apikey, verbose = 0)

p1 <- PlotOnStaticMap(denver_map, lat = denver_mj_licenses$LATITUDE, lon = denver_mj_licenses$LONGITUDE, destfile="denver_map.png", GRAYSCALE = FALSE, add = FALSE, FUN = points, NEWMAP = TRUE, TrueProj = TRUE, axes = TRUE, atX = NULL, atY = NULL, col = 3, pch=9)
legend("topleft", legend = "Location of Marijuana Businesses", col = "green", bg = "white", pch=9)

p2 <- PlotOnStaticMap(denver_map, lat = denver_mj_crime$LATITUDE, lon = denver_mj_crime$LONGITUDE, destfile="denver_map.png", GRAYSCALE = FALSE, add = FALSE, FUN = points, NEWMAP = TRUE, TrueProj = TRUE, axes = TRUE, atX = NULL, atY = NULL, col = 2, pch=12)
legend("topleft", legend = "Location of Marijuana Arrests", col = "red", bg = "white", pch=12)

```
